### 2019-03-07 experiments.tex
value pool formalism, parallel sort example

In order to evaluate our approach and implementation, we applied
nondeterminism detection to three real world examples.  The primary
points we wanted to explore were (1) whether our approach was able to reliably detect actual
nondeterminism, (2) whether the overhead of nondeterminism
detection for real systems was acceptable, and (3) the performance of
delta-debugging for real nondeterministic predicates.

@@

The {\tt redis-py} \cite{redispy} module implements a widely-used Python interface
to the popular Redis in-memory database, cache, and
message broker \cite{redis}.

@@

However, that a command such as {\tt restore} takes an expiration
argument is not obvious to a non-expert, nor is the behavior of {\tt
  spop} which pops a random value from a set.  Moreover, it is
difficult to guess whether the {\tt pipe} mechanism, which allows a
large sequence of commands to be queued up in a pipe and executed all
at once, introduced the potential for nondeterminism (does it execute
sequentially before other command are handled, or is it in parallel
with further commands).  Deep experience with Redis would make these
issues clear, but tests using a library are often written by those not
intimately familiar with the semantics of every used library (otherwise they would not have
been as likely to introduce flaky tests in the first place).  This is
even more true in the case of automated test generation, where the
test engineer is often chosen for expertise in test generation tools,
not in the domain of the software under test (SUT), and is seldom the
original developer of the code.

@@

Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
  datarray} algorithms.  The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.  Running this harness using TSTL's {\tt
  --checkProcessDeterminism} flag required less than 10 seconds on
average to produce a test exhibiting process-level nondeterminism in
the {\tt calc\_marginals\_sumproduct} function (the only broken
algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
showing an extremely simple input producing the issue, 
required another 92 seconds.  Interestingly, the way that {\tt
  python-afl} is used in TSTL means that AFL stability is 100\%,
because the {\tt PYTHONHASHSEED} has already been chosen before the
fork in AFL executions.

@@

The TSTL harness for {\tt pyfakefs} has been used to detect (and
correct) over 80 faults.  However, the testing largely relies on
the existence of a reference file system implementation.  One purpose
of failure nondeterminism is to make it somewhat easier to perform
property-based testing of
complex APIs like this even without a reference implementation, or in cases
where the implementations do not use the same error
codes (as is common, e.g. in NASA flight software
\cite{ICSEDiff,CFV08}).

>>>>

In order to evaluate our approach and implementation, we applied
nondeterminism detection to real Python code, including some widely
used systems ({\tt redis-py} and {\tt pyfakefs} are well-maintained,
widely-used libraries).  The primary
points we wanted to explore were (1) whether our approach was able to reliably detect actual
nondeterminism, (2) whether the overhead of nondeterminism
detection for real systems was acceptable, and (3) the performance of
delta-debugging for real nondeterministic predicates.

\subsection{Case Study: Parallel Sorting}

Our first case study is a simple implementation of a parallel merge
sort in Python
\url{https://gist.github.com/stephenmcd/39ded69946155930c347}.  The
merge sort and harness can be found in the TSTL GitHub repository
\cite{tstl}, under {\tt examples/parallelsorts}.  When a sorting
algorithm is implemented using concurrency, a key question to ask is
whether the ordering of elements that are equal under the comparison
operator is consistent across sorts of the same original sequence.
This is related to the question of whether a sort is a stable sort
(preserving the original relative order of equal-by-comparison elements),
but imposes a less strict requirement on the sort; a concurrent sort
might not be stable, but might still produce the same (non-stable)
ordering from the same provided sequence, every time.  This is a
particular instance of a common expectation noted above:  while a parallel
implementation of an algorithm will likely have internal
nondeterminism (due to scheduling of work), it is often expected to
behave deterministically, from the point of view of a caller.

This case study has two purposes:  first, to show that the
nondeterminism detection features of TSTL can be used to, with
acceptable overhead, show that a parallel sort implementation provides
a consistent, deterministic ordering, and second to show that the
mechanism can also detect the nondeterminism in a parallel sort that
does exhibit nondeterministic behavior.  For the second purpose, we
implemented a very simple swap-based parallel sorting algorithm that
repeatedly has multiple threads scan through a sequence and swap
out-of-order neighbors, until the sequence is sorted (a kind of
parallel bubble-sort with a simpler termination criteria).

\begin{figure}
{\scriptsize
\begin{code}
@import mergesort
@import swapsort
\vspace{0.1in}
pool: <val> 10
pool: <data> 10
pool: <sorted> 10
\vspace{0.1in}
<val> := <[-10..10]>
\vspace{0.1in}
<data> := []
<data>.append((<val>,<val>))
len(<data,2>) < len(<data,1>) -> <data>.extend(<data>)
\vspace{0.1in}
<sorted> := mergesort.merge\_sort\_parallel(<data>)
<sorted> := swapsort.swap\_sort\_parallel(<data>)
<sorted>
\vspace{0.1in}
property: swapsort.is\_sorted(<sorted>)
\end{code}
}
\caption {Complete TSTL harness for checking two parallel sorting
  algorithms for nondeterminism.}
\label{fig:parallelcode}
\end{figure}

Figure \ref{fig:parallelcode} shows the complete TSTL harness for
checking the parallel sorts.  This harness produces sequences of
integer pairs, where only the first integer is used in the comparison
operator of the sorting algorithms, and checks that the results of
{\tt sort\_parallel} calls are in fact sorted, using the {\tt
  swapsort}'s function for checking if a sequence is sorted.

@@

The {\tt redis-py} \cite{redispy} module implements a widely-used Python interface
to the popular Redis in-memory database, cache, and
message broker \cite{redis}.  The TSTL {\tt redis-py} harnesses can be found in
the {\tt examples/redis} directory in the TSTL GitHub repo \cite{tstl}.

@@

However, that a command such as {\tt restore} takes an expiration
argument is not obvious to a non-expert, nor is the behavior of {\tt
  spop} which pops a random value from a set.  Moreover, it is
difficult to guess whether the {\tt pipe} mechanism, which allows a
large sequence of commands to be queued up in a pipe and executed all
at once, introduced the potential for nondeterminism (does it execute
sequentially before other command are handled, or is it in parallel
with further commands).  Deep experience with Redis would make these
issues clear, but tests using a library are often written by those not
intimately familiar with the semantics of every library call (otherwise they would not have
been as likely to introduce flaky tests in the first place).  This is
even more true in the case of automated test generation, where the
test engineer is often chosen for expertise in test generation tools,
not in the domain of the software under test (SUT), and is seldom the
original developer of the code.

@@

Figure \ref{hashbug} shows TSTL code for generating inputs to the {\tt
  datarray} algorithms (it can be found in the {\tt
  examples/datarray\_inference} directory in the TSTL repository \cite{tstl}).  The {\tt flatten\_and\_sort} function is needed because we care about actual differences in probability values, not simply the order of list, tuple, or dictionary items.  Running this harness using TSTL's {\tt
  --checkProcessDeterminism} flag required less than 10 seconds on
average to produce a test exhibiting process-level nondeterminism in
the {\tt calc\_marginals\_sumproduct} function (the only broken
algorithm).  Reducing this 60 step test to a minimal test of only 6 steps,
showing an extremely simple input producing the issue, 
required another 92 seconds.  Interestingly, the way that {\tt
  python-afl} is used in TSTL means that AFL stability is 100\%,
because the {\tt PYTHONHASHSEED} has already been chosen before the
fork in AFL executions.

@@

The TSTL harness for {\tt pyfakefs} has been used to detect (and
correct) over 80 faults.  The harness can be found in the TSTL GitHub repository
in the {\tt examples/pyfakefs} directory, and the bugs discovered can
be viewed at \url{https://github.com/jmcgeheeiv/pyfakefs/issues?q=label%3ATSTL}.  However, the testing largely relies on
the existence of a reference file system implementation.  One purpose
of failure nondeterminism is to make it somewhat easier to perform effective
property-based testing of
complex APIs like this even without a complete reference implementation, or in cases
where the implementations do not use the same error
codes (as is common, e.g. in NASA flight software
\cite{ICSEDiff,CFV08}).

### 2019-03-07 implementation.tex
value pool formalism, parallel sort example

@@

In TSTL, a test consists of a sequence of
\emph{actions}, where actions execute arbitrary Python code, but are
expected to work by modifying the \emph{state} of a test, which is
stored in a fixed set of \emph{pools} containing Python objects.  The
TSTL actions defined by a test harness correspond to the set $A$ of
actions in Section \ref{sec:formal}, and transitions between states of
result from carrying out the Python code in an action.  Figure
\ref{fig:simple} shows a very simple TSTL harness that tests Python's
built-in {\tt set} and {\tt list} types.  The harness defines the
(visible) state of a test as consisting of three named pools of Python
objects, {\tt int}, {\tt l}, and {\tt s}.  These pools have no
inherent associated type, though a type can be annotated as an
additional constraint that TSTL will check holds during test
execution.  Initially all the pools hold the special Python value {\tt
  None} indicating they have not been initialized.  In terms of our formalism,
$I = \{${\tt [<int> = [None, None, None, None, None]}, {\tt <l> =
  [None, None, None, None, None]}, {\tt <s> = [None, None, None, None,
  None]]}$\}$.  Actions with the {\tt :=} assignment operator
initialize pools.  For example, the line of code {\tt l := []} defines a set of 5
actions, each of which initializes a different member of the list pool
{\tt l} to an empty list --- {\tt l0 = []} ... {\tt l4 = []}.  Actions
without a {\tt :=} require that pools used in them already be
initialized.  TSTL provides constructs like {\tt <[1..20]>} for
choosing values in a range for use in a test, and allows a user to
specify which exceptions an action may raise without causing the test
to fail (as in {\tt \{KeyError\} <s>.remove(<int>)}).

>>>>

\subsection{Pools and Visible State}

While the formal definitions of nondeterminism proposed in Section
\ref{sec:formal} offer a framework for checking for nondeterministic
behavior, they leave unspecified the key notions of \emph{state},
\emph{visible state}, and \emph{actions}.  For example, we could
identify the state and visible state of a system as identical with the
full set of memory locations accessed by that system, and actions with the
individual processor instructions executed.  However, this
definition is highly impractical for two reasons.  First, it is
extremely inefficient to compare multiple executions for
full-memory-state equivalence at every instruction step.  Second, and
even more importantly, as noted in the introduction, very few programs
will behave deterministically when considered at this low level,
unless the full state of the operating system, processor, hardware
devices, etc. are all taken as controlled inputs.

What is a practical level of granularity for (visible) states and actions that
can be efficiently checked, and matches (for the most part) developer
and tester intuitions about what should behave deterministically?

Andrews et al. formalized a value-pool-based model of unit tests that is widely used in
actual testing tools, including Randoop \cite{Pacheco}, showing that
all Java unit tests (and, more generally, unit tests in similar
languages) can be represented in a \emph{canonical} form, where
a test consists of three parts:

\begin{enumerate}
\item A declaration of a set of array variables, referred to as
  \emph{value pools}, e.g., {\tt int [] intVP = new int[3];}.
  \item A set of assignments of constant values to elements of
    primitive type value pools, e.g. {\tt intVP[2] = 1;}.
    \item A part in which all statements are assignments of calls of a
      method to a value pool, with all arguments to the method also
      taken from a value pool, e.g. {\tt intVP[2] = fooObjectVP[2].bar(intVP[0],
        intVP[1]);} (these are referred to as \emph{array-canonical} statements).
    \end{enumerate}

Given such a form for unit tests, there is an obvious mapping to
states and actions in our formalism for nondeterminism:  the state is
the full state of the system, but visible state is restricted to
\emph{the values stored in the value pools}.  Actions are the
array-canonical statements that call code under test to modify the pools.  We can ignore constant assignments and
declarations, since (assuming the language implementation itself is
basically deterministic) these cannot introduce nondeterminism.
Because assertions of correctness and \emph{what a test actually does}
depend on the values in value pools, it seems likely that developers
and testers expect determinism in unit tests, at the granularity of
value pool assignments.

@@

In TSTL, a test consists of a sequence of
\emph{actions}, where actions execute arbitrary Python code, but are
expected to work by modifying the \emph{state} of a test, which is
stored in a fixed set of pools containing Python objects.  TSTL's
approach essentially follows the pool-based canonical form defined by
Andrews et al., but relaxes many of its restrictions
(e.g., a constant value can be used in a method call without first
assigning it to a pool). The
TSTL actions defined by a test harness correspond to the set $A$ of
actions in Section \ref{sec:formal}, and transitions between states
result from executing out the Python code in an action; these actions
are, roughly, the array-canonical statements of the value pool formalism.  Figure
\ref{fig:simple} shows a very simple TSTL harness that tests Python's
built-in {\tt set} and {\tt list} types.  The harness defines the
(visible) state of a test as consisting of three named pools of Python
objects, {\tt int}, {\tt l}, and {\tt s}.  These pools have no
inherent associated type, though a type can be annotated as an
additional constraint that TSTL will check holds during test
execution.  Initially all the pools hold the special Python value {\tt
  None} indicating they have not been initialized.  In terms of our formalism,
$I = \{${\tt [<int> = [None, None, None, None, None]}, {\tt <l> =
  [None, None, None, None, None]}, {\tt <s> = [None, None, None, None,
  None]]}$\}$.  Actions with the {\tt :=} assignment operator
initialize pools.  For example, the line of code {\tt l := []} defines a set of 5
actions, each of which initializes a different member of the list pool
{\tt l} to an empty list --- {\tt l0 = []} ... {\tt l4 = []}.  Actions
without a {\tt :=} require that pools used in them already be
initialized.  TSTL provides constructs like {\tt <[1..20]>} for
choosing values in a range for use in a test, and allows a user to
specify which exceptions an action may raise without causing the test
to fail (as in {\tt \{KeyError\} <s>.remove(<int>)}).
